BOLD Parameter Estimation Using Particle Filters

### 2010-07-19 doc/thesis/background.tex
more work....
mostly done, just a bit of cleanup

Currently, FMRI is used to determine the location of responses
to stimuli. The method of finding activation is discussed in 
\autoref{sec:Statistical Parameteric Mapping}. The goal of this 
work is to move away from the question "Is this region active"
and instead ask "What does the activation look like". Answering
this question necessitates more complex models and certainly
will result in longer run times. Already there have been several other
attempts to model the BOLD response; these works will be discussed
in [all other sections basically].

\section{Statistical Parametric Mapping}
Although not strictly the same as parameter calculation from 
FMRI, activation detection is similar and worth discussing. Estimation of 
parameters is a generalization of the idea of activation detection.
Given the popularity of Statistical Parametric Mapping (SPM) 
it is important to draw a distinction between the methods proposed
in this work.

@@

\subsection{Conclusion}
In all, the GLM is extremely useful for determining linear 
dependence of a set of regressors on the output. Unfortunately, as discussed in
\autoref{sec:BOLD Analysis} there are significant nonlinearities 
that almost certainly cause false negatives in the Statistical Parametric
Maps. Unfortunately nonlinear analyses have only recently become feasible,
so the scale of the problem is still unknown. The problem is 
highlighted by the relatively common scenario where no significant
activation can be found in a single run \cite{Riera2004}
\cite{Johnston2008}.

@@

[simulated annealing image?]

@@

[genetic algorithm picture]

@@

\begin{figure}
\caption{Examples of update distributions, using an Kalman Filter, \cite{Thrun2005}}
\label{fig:EKFWorking}
\end{figure}

>>>>

Currently, FMRI is used to determine the location of responses
to stimuli. The method of finding activation is discussed in 
\autoref{sec:Statistical Parametric Mapping}. The goal of this 
work is to move away from the question "Is this region active"
and instead ask "What does the activation look like". Answering
this question necessitates more complex models and certainly
will result in longer run times. Already there have been several other
attempts to model the BOLD response; these works will be discussed
in in the next few sections.

\section{Statistical Parametric Mapping}
\label{sec:Statistical Parametric Mapping}
Although not strictly the same as parameter calculation from 
FMRI, activation detection is similar and worth discussing. Estimation of 
parameters is a generalization of the idea of activation detection.
Given the popularity of Statistical Parametric Mapping (SPM) 
it is important to draw a distinction between the methods proposed
in this work.

@@

\subsection{Conclusion}
\label{sec:BackgroundConclusion}
In all, the GLM is extremely useful for determining linear 
dependence of a set of regressors on the output. Unfortunately, as discussed in
\autoref{sec:BOLD Analysis} there are significant nonlinearities 
that almost certainly cause false negatives in the Statistical Parametric
Maps. Unfortunately nonlinear analyses have only recently become feasible,
so the scale of the problem is still unknown. The problem is 
highlighted by the relatively common scenario where no significant
activation can be found in a single run \cite{Riera2004}
\cite{Johnston2008}.

@@

%[simulated annealing image?] todo

@@

%[genetic algorithm picture] todo

@@

\begin{figure}
\includegraphics[width=16cm]{images/kalman}
%\includegraphics[trim=6cm .75cm 6cm .75cm,width=16cm]{images/gauss_step_point1sec_3sigma.pdf}
\caption{Example updates of a distribution using Kalman Filter, \cite{Thrun2005}}
\label{fig:EKFWorking}
\end{figure}

### 2010-07-19 doc/thesis/conclusion.tex
mostly done, just a bit of cleanup

\chapter{Conclusion}

\section{Future Work}
*A more extensive investigation of the parameters distribution
*Using a compliance model
*restrict parameters more (may not be possible)

>>>>

\chapter{Conclusion}
In this paper I have proposed and demonstrated the use of a particle filter
for the estimation of BOLD model parameters. In doing so, it has been possible
to estimate the BOLD time-series and judge the quality of the estimated time-
series with respect to the true BOLD signal. As result, it is possible to use
the particle filter to localize areas where a known stimuli most directly drives
neural activation. In the future it will be possible to use the estimated states
($v,q,s,f$) to drive other models and learn more about how regions of the brain
interact.

A limitation often reached in previous works was an inconsistency of
parameter estimates, most likely because of covariance between the model parameters.
Although individual studies got consistent results, those results often differed
widely from other similar studies. The reason for this is rather clear from the
simulation results in \autoref{sec:SimLowNoise}. There is a significant amount of
trade off between parameters to the point that a signal set of parameters
is most likely not possible to derive from the BOLD response alone. It
will therefore be beneficial to combine BOLD studies with cerebral blood
flow or cerebral blood volume studies to gain multiple more measurements and
further constrain the model. That said, the benefit of the particle filter
is that it provides a full posterior distribution at the final time step.
As such the true solution should be encoded in the particle filter's final 
distribution given priors that encompass the true parameters. This is beneficial
in two ways; first, if, after the fact, some parameter becomes known 
from outside observation, it is then possible to construct a new probability
conditional on the new observation. Secondly the results from multiple runs
may be reasonably compiled together, using the final distribution from the previous
run as the prior distribution of the next run. Rather than simply providing a 
staring point for parameters to converge from, it in fact continues convergence
from the previous stopping point.

Although many versions of the BOLD model exist, and it is tempting to use 
more detailed models; from the results found here the issue of bias error
from the BOLD model is not the biggest concern. Clearly reducing variance
in parameters should be the first concern. 
One future study that would be extremely beneficial in this way  would
be an extensive study of what the priors should truly be. Although \cite{Friston2000}
gives an estimate of what is thought to be "reasonable" values, and later studies
published their estimate of the distributions, given the interplay between
parameters it is very unlikely that these priors are true to actual distribution
that occurs in vivo. As I mentioned previously, the addition of simultaneous
flow or volume measurements are another potentially powerful way to further confine the model,
and thus deal with the elasticity of parameters. As I mentioned in 
\autoref{sec:BackgroundConclusion}, a chief advantage of using physiologically
plausible models is that such data may in fact be added with relative ease.

In conclusion, using particle filters to estimate the BOLD response are 
an extremely powerful method of fitting to noisy data. The technique also
holds great promise as extensible platform to build more advanced models
and techniques on top of. Integrating further information is necessary to
move beyond the traditional Statistical Parametric Mapping and moving
toward biologically and medically relevant FMRI scanning techniques.

### 2010-07-19 doc/thesis/intro.tex
added quite a few plots. Should be able to write results for single voxel simulations.
mostly done, just a bit of cleanup

\section{Noise}
\label{sec:Introduction Noise}
As demonstrated in \autoref{sec:BOLD Phsyiology} the BOLD response has been
extensively studied and despite minor discrepancies, the cause of the BOLD 
signal is well known. However, as FMRI detects an  
aggregate signal over the space of cubic centimeters, there are
plenty of sources of noise. Though local neurons act
"together" (i.e. around the same time), the density of neurons, the
density of capillaries, and slight differences in activation across 
a particular voxel can all lead to signal attenuation and noise.

@@

In order to characterize the noise, I analyzed resting state data.
During resting state, the patient is shown no images, and he is asked
to avoid movement and complex thought, to the best of his abilities.
Some limitations exist, for instance EPI sequences are extremely loud
which could cause some stimulus in the auditory cortex, and of course
very long sequences could give the patient time for his mind to wander.
Overall though there should be no activation, and thus the signal will
consist entirely of noise. Therefore resting state data is perfect
for analyzing the noise distribution. The locations were chosen from
various points around the brain, all in grey matter voxels (see \autoref{sec:methods}
for a discussion on how grey matter voxels were found). The time
series were also chosen because they were representative of different types
of noise I found in the resting state data.

Because most methods (including the one used in this paper)
assume the noise realizations are independent of each other, the auto-
correlation is of particular interest (which is a necessary but not
sufficient condition for independence). Gaussianity is also a common
assumption made in studies of FMRI data, though that assumption is not
made in this work. Regardless, comparing the distribution to a Gaussian
is informative, so I used Q-Q plots to compare example data with the
Normal distribution. Additionally, in FMRI data the noise is often considered 
to be Wiener \cite{Riera2003}. Recall that a Wiener random process is
characterized by steps that are Gaussian; in other words the difference between
any two samples is Normally distributed. The simulations discussed in 
\autoref{sec:Methods Single Voxel Simulation} make use of this, 
by adding a Wiener random process to the overall signal. To determine
whether the noise is in fact Wiener, the distribution of 
the difference between adjacent measurements was plotted against
a Gaussian. Wiener processes further assume that the steps are independent;
so the steps should have near flat autocorrelation functions.

@@

\begin{table}[t]
\centering
\begin{tabular}{|c || c | c | c | c|}
\hline 
Parameter  & \cite{Friston2002} & \cite{Johnston2008} & \cite{Vakorin2007} & \cite{Deneux2006}\\
\hline
$\epsilon$ &  $N(.54 , .1 ^2)$  & $.069 \pm .014 $ & (NC) & .89\\
$\tau_s  $ &  $N(1.54, .25^2)$  & $4.98 \pm 1.07 $ & $2.2$ & 2.04 \\
$\tau_f  $ &  $N(2.46, .25^2)$  & $8.31 \pm 1.51 $ & $.45$ & 5.26\\
$\tau_0  $ &  $N(.98 , .25^2)$  & $8.38 \pm 1.5  $ & $.94$ & .27\\
$\alpha  $ &  $N(.33 , .45^2)$  & $.189 \pm .004 $ & $.4$ (NC) & .63 \\
$E_0     $ &  $N(.34 , .1 ^2)$  & $.635 \pm .072 $ & $.6$ (NC) & .33\\
$V_0     $ &  $.03$ (NC)        & $.0149 \pm .006$ & (NC) & .16\\
\hline
\end{tabular}
\caption{Parameters found by various studies. (NC) indicates that the value
wasn't calculated. \cite{Vakorin2007} made use of the values from \cite{Friston2002}
where not explicitly stated}
\label{tab:Params} 
\end{table}

>>>>

\section{Noise}
\label{sec:Introduction Noise}
As demonstrated in \autoref{sec:BOLD Physiology} the BOLD response has been
extensively studied and despite minor discrepancies, the cause of the BOLD 
signal is well known. However, as FMRI detects an  
aggregate signal over the space of cubic centimeters, there are
plenty of sources of noise. Though local neurons act
"together" (i.e. around the same time), the density of neurons, the
density of capillaries, and slight differences in activation across 
a particular voxel can all lead to signal attenuation and noise.

@@

In order to characterize the noise, I analyzed resting state data.
During resting state, the patient is shown no images, and he is asked
to avoid movement and complex thought, to the best of his abilities.
Some limitations exist, for instance EPI sequences are extremely loud
which could cause some stimulus in the auditory cortex, and of course
very long sequences could give the patient time for his mind to wander.
Overall though there should be no activation, and thus the signal will
consist entirely of noise. Therefore resting state data is perfect
for analyzing the noise distribution. The locations were chosen from
various points around the brain, all in grey matter voxels. The time
series were also chosen because they were representative of different types
of noise I found in the resting state data.

Because most methods (including the one used in this paper)
assume the noise realizations are independent of each other, the auto-
correlation is of particular interest (which is a necessary but not
sufficient condition for independence). Gaussianity is also a common
assumption made in studies of FMRI data, though that assumption is not
made in this work. Regardless, comparing the distribution to a Gaussian
is informative, so I used Q-Q plots to compare example data with the
Normal distribution. Additionally, in FMRI data the noise is often considered 
to be Wiener \cite{Riera2003}. Recall that a Wiener random process is
characterized by steps that are Gaussian; in other words the difference between
any two samples is Normally distributed. The simulations discussed in 
\autoref{sec:Single Voxel Simulation} make use of this, 
by adding a Wiener random process to the overall signal. To determine
whether the noise is in fact Wiener, the distribution of 
the difference between adjacent measurements was plotted against
a Gaussian. Wiener processes further assume that the steps are independent;
so the steps should have near flat autocorrelation functions.

@@

\begin{table}[t]
\centering
\begin{tabular}{|c || c | c | c | c|}
\hline 
Parameter  & \cite{Friston2000} & \cite{Johnston2008} & \cite{Vakorin2007} & \cite{Deneux2006}\\
\hline
$\tau_0  $ &  $N(.98 , .25^2)$  & $8.38 \pm 1.5  $ & $.94$ & .27\\
$\alpha  $ &  $N(.33 , .45^2)$  & $.189 \pm .004 $ & $.4$ (NC) & .63 \\
$E_0     $ &  $N(.34 , .1 ^2)$  & $.635 \pm .072 $ & $.6$ (NC) & .33\\
$V_0     $ &  $.03$ (NC)        & $.0149 \pm .006$ & (NC) & .16\\
$\tau_s  $ &  $N(1.54, .25^2)$  & $4.98 \pm 1.07 $ & $2.2$ & 2.04 \\
$\tau_f  $ &  $N(2.46, .25^2)$  & $8.31 \pm 1.51 $ & $.45$ & 5.26\\
$\epsilon$ &  $N(.54 , .1 ^2)$  & $.069 \pm .014 $ & (NC) & .89\\
\hline
\end{tabular}
\caption{Parameters found by various studies. (NC) indicates that the value
wasn't calculated. \cite{Vakorin2007} made use of the values from \cite{Friston2002}
where not explicitly stated}
\label{tab:Params} 
\end{table}
